### 2023-07-31 proofs/main.tex
wip
wip: leftovers

\section{The Vanilla System}

First, we present the top-level system, which is easy to understand.

\subsection{Grammar}
\ottgrammartabular{
  \ottP\ottinterrule
  \ottN\ottinterrule
  \ottv\ottinterrule
  \ottc\ottinterrule
}

@@

\subsection{Declarative Subtyping}
\ottdefnsDZero

\section{Multi-Quantified System}
\subsection{Grammar}
\ottgrammartabular{
  \ottiP\ottinterrule
  \ottiN\ottinterrule
}
\subsection{Declarative Multiquantified Subtyping}
\ottdefnsDOne

@@

\subsection{Constraint Merge}

Unification and subtyping constraints is are by a list of constraint entries.
Each entry restricts an unification variable in two possible ways: either
stating that it must be equivalent to a certain type ($[[pua :≈ iP]]$ or $[[nua
:≈ iN]]$) or that it must be a (positive) supertype of a certain type ($[[pua :≥ iP]]$).

\begin{definition} [Matching Entries]
  We call two entries matching if they are restricting the same unification variable.
\end{definition}

Two matching entries can be merged in the following way:
\begin{definition} \hfill \\
\ottdefnSCME\\
\end{definition}
% Notice that in case of equivalence, the assigned types
% must be equal (i.e. alpha-equivalent) to be merged. This is because
% the unification algorithm assumes that every type is normalized,
% and hence, equivalence is alpha-equivalence 
% (\cref{corollary:nf-complete-wrt-subt-equiv,corollary:nf-sound-wrt-subt-equiv}).

To merge two constraints, we merge each pair of
matching entries, and unite the results.
\ilyam{add contexts}
\begin{definition}
  $\begin{aligned}[t] [[SC1 & SC2]] &= 
    \{ [[scE1 & scE2]] \mid [[scE1]] \in [[SC1]], [[scE2]]  \in [[SC2]], \text{s.t. } [[scE1]] \text{ matches with } [[scE2]] \}\\
    &\cup 
    \{ [[scE1]] \mid [[scE1]] \in [[SC1]], \text{ s.t. } \forall [[scE2]]  \in [[SC2]], [[scE1]] \text{ does not match with } [[scE2]] \}\\
    &\cup 
    \{ [[scE2]] \mid [[scE2]] \in [[SC2]], \text{ s.t. } \forall [[scE1]]  \in [[SC1]], [[scE1]] \text{ does not match with } [[scE2]] \}\\
   \end{aligned}$
\end{definition}

@@

\subsection{Typing}
\input{../_build/_gen/tex/alg-typing-lemmas.tex}

>>>>

\tableofcontents

\newpage

\section{Declarative System}
First, we present the top-level system, which is easy to understand.

\subsection{Grammar}
\ottgrammartabular{
  \ottiP\ottinterrule
  \ottiN\ottinterrule
  \ottv\ottinterrule
  \ottc\ottinterrule
}

@@

\subsection{Declarative Subtyping}
\ottdefnsDOne

@@

\subsection{Constraints}

Unification and subtyping algorithms are based on the constraint generation.
The constraints are represented by set of constraint entries.

\begin{definition}[Unification Constraint]
  \hfill
  \begin{itemize}
    \item Unification entry (denoted as $[[ucE]]$) is an expression of shape 
      $[[pua :≈ iP]]$ or $[[nua :≈ iN]]$;
    \item unification constraint (denoted as $[[UC]]$) is a set of 
      unification constraint entries.
  \end{itemize}

\end{definition}

\begin{definition}[Subtyping Constraint]
  \hfill
  \begin{itemize}
    \item Subtyping entry (denoted as $[[scE]]$) is an expression of shape 
      $[[pua :≥ iP]]$, $[[nua :≈ iN]]$, or $[[pua :≈ iP]]$;
    \item subtyping constraint (denoted as $[[SC]]$) is a set of subtyping constraint entries.
  \end{itemize}
\end{definition}

\begin{definition}[Well-formed Constraint Entry]
  We say that a constraint entry is well-formed in a context $[[Γ]]$ if
  the type it restricts the unification variable to is well-formed in $[[Γ]]$.
  \begin{itemize}
    \item $[[Γ ⊢ pua :≥ iP]]$ iff $[[Γ ⊢ iP]]$;
    \item $[[Γ ⊢ pua :≈ iP]]$ iff $[[Γ ⊢ iP]]$;
    \item $[[Γ ⊢ nua :≈ iN]]$ iff $[[Γ ⊢ iN]]$.
  \end{itemize}
\end{definition}

\begin{definition} [Matching Entries]
  We call two unification constraint entries 
  or two subtyping constraint entries matching 
  if they are restricting the same unification variable.
\end{definition}

Two matching entries formed in the same context $[[Γ]]$ 
can be merged in the following way:
\begin{definition}[Merge of Matching Constraint Entries]
   \hfill \\
\ottdefnSCME\\
\end{definition}
% Notice that in case of equivalence, the assigned types
% must be equal (i.e. alpha-equivalent) to be merged. This is because
% the unification algorithm assumes that every type is normalized,
% and hence, equivalence is alpha-equivalence 
% (\cref{corollary:nf-complete-wrt-subt-equiv,corollary:nf-sound-wrt-subt-equiv}).

\begin{definition}[Constraint Context]
  Constraint context (denoted as $[[Θ]]$) is a set of entries of shape $[[ pua[Γ] ]]$ 
  and $[[ nua[Γ] ]]$, specifying that the unification variable ($[[pua]]$ or $[[nua]]$)
  must be instantiated in a type well-formed in context $[[Γ]]$.
  We assume that for each $[[α̂±]]$, at most one entry of shape $[[ α̂±[Γ] ]]$ is in $[[Θ]]$.
  If $[[ α̂±[Γ] ]] \in [[Θ]]$, we denote $[[Γ]]$ as $[[Θ(α̂±)]]$.
\end{definition}

\begin{definition}[Well-formed Constraint]
  We say that a constraint is well-formed in a
  constraint context $[[Θ]]$ if all its entries are well-formed in
  the corresponding elements of $[[Θ]]$.
  More formally, 
  $[[Θ ⊢ UC]]$ iff for every $[[ucE]] \in $ $[[UC]]$,
  such that $[[ucE]]$ restricts $[[α̂±]]$,
  there exists $[[ α̂±[Γ] ]] \in [[Θ]]$
  and $[[Γ ⊢ ucE]]$.
\end{definition}

\begin{definition}[Merge of Subtyping Constraints]
  Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$.
  Then $[[Θ ⊢ SC1 & SC2 = SC]]$
  defines a set such that $[[ucE]] \in [[SC]]$ iff either
  \begin{itemize}
    \item $[[ucE]] \in [[SC1]]$ and there is no matching $[[ucE']] \in [[SC2]]$; or
    \item $[[ucE]] \in [[SC2]]$ and there is no matching $[[ucE']] \in [[SC1]]$; or
    \item $[[Θ(α̂±) ⊢ ucE1 & ucE2 = ucE]]$ for some $[[ucE1]] \in [[SC1]]$ and $[[ucE2]] \in [[SC2]]$
      such that $[[ucE1]]$ matches with $[[ucE2]]$ restricting variable
      $[[α̂±]]$. 
  \end{itemize}
\end{definition}

@@

\subsection{Declarative Typing}
\input{../_build/_gen/tex/decl-typing-lemmas.tex}

\subsection{Algorithmic Typing}
\input{../_build/_gen/tex/alg-typing-lemmas.tex}

### 2023-07-31 proofs/tex/alg-pos-subtyping-lemmas.tex
wip: leftovers

\begin{lemma}[Completeness of the Positive Subtyping] \label{lemma:pos-subt-completeness}
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; Θ ⊢ uP]]$.
    Then for any $[[Θ ⊢ uσ]]$ such that $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
    there exists $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ such that $[[Θ ⊢ SC]]$ and moreover, $[[Θ ⊢ uσ : SC]]$.
\end{lemma}
\begin{proof}
    Let us prove this lemma by induction on $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$.
    Let us consider the last rule used in the derivation,
    but first, consider the base case for the substitution $[[ [uσ]uP ]]$:
    \begin{caseof}
        \item \label{case:pos-subt-complete-base} $[[uP]] = [[ ∃nbs.α̂⁺ ]]$ (for potentially empty $[[nbs]]$)\\
        Then by assumption, $[[ Γ ⊢ ∃nbs.[uσ]α̂⁺ ≥ iQ ]]$ (where $[[ {nbs} ∩ fv [uσ]α̂⁺ = ∅]]$).
        By \cref{todo}, $[[ Γ ⊢ ∃nbs.[uσ]α̂⁺ ≥ iQ ]]$ means 
        $[[ Γ ⊢ [iNs/nbs][uσ]α̂⁺ ≥ iQ ]]$, and hence $[[ Γ ⊢ [uσ]α̂⁺ ≥ iQ ]]$.
        By inversion, $[[Γ ; Θ ⊢ ∃nbs.α̂⁺]]$ implies $[[â⁺[Δ] ∊ Θ]]$ for some $[[Δ]]$.

        In the algorithm trying to infer the subtyping 
        $[[Γ; Θ ⊨ ∃nbs.α̂⁺ ≥ iQ ⫤ SC]]$,
        after multiple applications of \ruleref{\ottdruleAExistsLabel}
        the type $[[∃nbs.α̂⁺]]$ is reduced to $[[α̂⁺]]$.
        Next, the algorithm tries to apply
        \ruleref{\ottdruleAPUVarLabel}
        and the resulting restriction is $[[SC']] = [[(α̂⁺ :≥ iQ')]]$ where
        $[[upgrade Γ ⊢ iQ to Δ = iQ']]$.

        Why does the upgrade procedure terminates?
        Because $[[ [uσ]α̂⁺ ]]$ satisfies the pre-conditions of the completeness of the upgrade
        (\cref{lemma:upgrade-completeness}):
        \begin{itemize}
            \item $[[Δ ⊢ iP']]$ because $[[iP' = [uσ]α̂⁺]]$ and $[[Θ ⊢ uσ]]$ and 
            $[[α̂⁺[Δ] ∊ Θ]]$,
            \item $[[Γ ⊢ iP' ≥ iQ]]$ as noted above
        \end{itemize}
        Moreover, the completeness of the upgrade also gives us $[[Γ ⊢ iP' ≥ iQ']]$
        and further, we strengthen it to $[[Δ ⊢ iP' ≥ iQ']]$
        (since by the soundness of the upgrade (\cref{lemma:upgrade-soundness}),
        $[[Δ ⊢ iQ']]$).  It means that $[[Δ ⊢ iP' : (α̂⁺ :≥ iQ') ]]$, that is 
        $[[Θ ⊢ uσ : (α̂⁺ :≥ iQ') ]]$, as required.

        \item \label{case:pos-subt-complete-pvar}
        $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ is derived by \ruleref{\ottdruleDOnePVarLabel}, 
        i.e. $[[iP]] = [[ [uσ]uP ]] = [[ α⁺ ]] = [[iQ]]$.
        Here the first equality holds because $[[uP]]$ is not a unification variable:
        it has been covered by \cref{case:pos-subt-complete-base}.
        The second equality hold because \ruleref{\ottdruleDOnePVarLabel} was applied.

>>>>

\begin{lemma}[Completeness of the Positive Subtyping] \label{lemma:pos-subt-completeness}
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; Θ ⊢ uP]]$.
    Then for any $[[Θ ⊢ uσ]]$ such that $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
    there exists $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ and moreover, $[[Θ ⊢ uσ : SC]]$.
\end{lemma}
\begin{proof}
    Let us prove this lemma by induction on $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$.
    Let us consider the last rule used in the derivation,
    but first, consider the base case for the substitution $[[ [uσ]uP ]]$:
    \begin{caseof}
        \item \label{case:pos-subt-complete-base} $[[uP]] = [[ ∃nbs.α̂⁺ ]]$ 
            (for potentially empty $[[nbs]]$)\\
            Then by assumption, $[[ Γ ⊢ ∃nbs.[uσ]α̂⁺ ≥ iQ ]]$ (where $[[ {nbs} ∩ fv [uσ]α̂⁺ = ∅]]$).
            Let us decompose $[[iQ]]$ as $[[iQ]] = [[∃ncs.iQ0]]$, where $[[iQ0]]$ does
            not start with $[[∃]]$.

            By inversion, $[[Γ ; Θ ⊢ ∃nbs.α̂⁺]]$ implies $[[â⁺[Δ] ∊ Θ]]$ for some 
            $[[{Δ} ⊆ {Γ}]]$.

            By \cref{lemma:quant-rule-decomposition} applied twice, 
            $[[ Γ ⊢ ∃nbs.[uσ]α̂⁺ ≥ ∃ncs.iQ0 ]]$ implies
            $[[ Γ,ncs  ⊢ [iNs/nbs][uσ]α̂⁺ ≥ iQ0 ]]$ for some $[[iN]]$, 
            and since $[[ {nbs} ∩ fv([uσ]α̂⁺) ⊆ {nbs} ∩ {Θ(α̂⁺)} ⊆ {nbs} ∩ {Γ} = ∅ ]]$,
            $[[ [iNs/nbs][uσ]α̂⁺ = [uσ]α̂⁺ ]]$, that is $[[ Γ,ncs ⊢ [uσ]α̂⁺ ≥ iQ0]]$.

            When algorithm tires to infer the subtyping 
            $[[Γ; Θ ⊨ ∃nbs.α̂⁺ ≥ ∃ncs.iQ0 ⫤ SC]]$,
            it applies \ruleref{\ottdruleAExistsLabel},
            which reduces the problem to
            $[[Γ, ncs; Θ, β̂⁻*[Γ, ncs] ⊨ [β̂⁻*/nbs] α̂⁺ ≥ iQ0 ⫤ SC]]$, 
            which is equivalent to 
            $[[Γ, ncs; Θ, β̂⁻*[Γ, ncs] ⊨ α̂⁺ ≥ iQ0 ⫤ SC]]$.

            Next, the algorithm tries to apply
            \ruleref{\ottdruleAPUVarLabel}
            and the resulting restriction is $[[SC]] = [[(α̂⁺ :≥ iQ0')]]$ where
            $[[upgrade Γ, ncs ⊢ iQ0 to Δ = iQ0']]$.

            Why does the upgrade procedure terminate?
            Because $[[ [uσ]α̂⁺ ]]$ satisfies the pre-conditions of the completeness of the upgrade
            (\cref{lemma:upgrade-completeness}):
            \begin{enumerate}
                \item $[[Δ ⊢ [uσ]α̂⁺ ]]$ because $[[Θ ⊢ uσ]]$ and $[[α̂⁺[Δ] ∊ Θ]]$,
                \item $[[ Γ,ncs ⊢ [uσ]α̂⁺ ≥ iQ0]]$ as noted above
            \end{enumerate}

            Moreover, the completeness of upgrade also says that $[[iQ0']]$ is 
            \emph{the least} supertype of $[[iQ0]]$ among types well-formed in $[[Δ]]$, 
            that is $[[Δ ⊢ [uσ]α̂⁺ ≥ iQ0']]$, which means 
            $[[Θ ⊢ uσ : (α̂⁺ :≥ iQ0')]]$, that is $[[Θ ⊢ uσ : SC]]$.

        \item \label{case:pos-subt-complete-pvar}
        $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ is derived by \ruleref{\ottdruleDOnePVarLabel}\\
        Then $[[iP]] = [[ [uσ]uP ]] = [[ α⁺ ]] = [[iQ]]$, where
        the first equality holds because $[[uP]]$ is not a unification variable:
        it has been covered by \cref{case:pos-subt-complete-base}; and
        the second equality hold because \ruleref{\ottdruleDOnePVarLabel} was applied.
